---
title: "Interactive R and more..."
output: 
  rmdformats::material
editor_options: 
  chunk_output_type: console
header-includes:
   - \usepackage{amssymb}
   - \usepackage{amsmath}
---

# Interactive R with RMarkdown

This document demonstrates how to use various R packages to add interactivity and utilize different programming languages within RStudio. You'll learn about popular packages like `leaflet` and `plotly`, which enable interactive data visualizations, as well as how to incorporate languages like Python, C++, and Bash directly into RMarkdown.

For source code see [github page](https://github.com/strnda/PED/blob/main/R/05_markdown.Rmd).

# Package Setup

```{r, message = FALSE, warning = FALSE, cache = FALSE}
lop <- c("reticulate", "Rcpp", "leaflet", "ggplot2", "plotly", "visNetwork", "DT")

to_instal <- lop[which(x = !lop %in% installed.packages()[,"Package"])]

if(length(x = to_instal) != 0) {
  install.packages(pkgs = to_instal)
}

aux <- lapply(X = lop, FUN = library, character.only = TRUE)

rm(list = c("lop", "to_instal", "aux"))
gc()
```

This code block installs and loads a set of essential packages used in this document. Here’s a quick overview of each:

- **`reticulate`**: Integrates Python into R, allowing you to run Python code in R scripts.
- **`Rcpp`**: Links R with C++ code, enabling high-performance computation.
- **`leaflet`**: Used for creating interactive maps.
- **`ggplot2`**: A powerful data visualization package.
- **`plotly`**: Provides interactive plots and works seamlessly with `ggplot2`.
- **`visNetwork`**: Allows creation of network visualizations.
- **`DT`**: Displays data tables with interactive features.

# Integrating Other Languages in RStudio {.tabset}

RStudio supports multi-language workflows, letting you run code in Python, C++, and Bash within an RMarkdown file.

## Python

Using the `reticulate` package, we can run Python code in RStudio.

### Example: Using Python Variables in R

In this example, we create a variable in Python and access it from R.

```{python}
x = .1 + .1 + .1
print(x)
```

```{r}
py$x
```

### Example: Using R Variables in Python

Similarly, we can define a variable in R and access it in Python.

```{r}
x <- rnorm(n = 10)
```

```{python}
print(r.x)
```

## C++

With the `Rcpp` package, we can write C++ code directly in RMarkdown. Here’s a simple example function written in C++ to double each element of a vector.

```{Rcpp}
#include <Rcpp.h>

//[[Rcpp::export]]
Rcpp::IntegerVector add_number(Rcpp::IntegerVector x) {
  return x + x;
}
```

We can call this function in R:

```{r}
add_number(c(2, 184))
```

## Bash

Bash commands can be executed in RMarkdown, making it easy to interact with the command line. For more Bash examples, see [this link](https://github.com/strnda/tools/blob/master/txt/linux).

```{bash}
echo "Hello Bash!"
ls -la
```

# Interactive HTML Widgets {.tabset}

HTML widgets in R enable interactive data visualizations directly within RMarkdown.

## Leaflet: Interactive Maps

`leaflet` is a powerful package for creating interactive maps in R. In this example, we use earthquake data (`quakes`) and map the locations with markers.

```{r}
data(quakes)
head(quakes)

leaflet(data = quakes[1:20,]) %>%
  addTiles() %>%
  addMarkers(lng = ~long, lat = ~lat, popup = ~as.character(x = mag), label = ~as.character(x = mag))
```

## Plotly: Interactive Graphs {.tabset}

The `plotly` package allows you to create interactive plots. It integrates well with `ggplot2`, allowing static `ggplot` objects to be made interactive.

### ggplot2 with Plotly

Here’s a bar plot of diamond cuts by clarity, enhanced with `plotly`.

```{r}
p <- ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) +
            geom_bar(position = "dodge")

ggplotly(p = p)
```

### Pure Plotly

You can also use `plotly` directly without `ggplot2` for custom visualizations.

```{r}
plot_ly(z = ~volcano, type = "heatmap")

plot_ly(z = ~volcano, type = "surface")
```

## visNetwork: Network Visualizations

The `visNetwork` package enables interactive network visualization in R.

```{r}
nodes <- data.frame(id = 1:6, title = paste("node", 1:6), shape = c("dot", "square"), size = 10:15, color = c("blue", "red"))
edges <- data.frame(from = 1:5, to = c(5, 4, 6, 3, 3))
head(x = nodes)

visNetwork(nodes = nodes, edges = edges)
```

The following example demonstrates customization with different node shapes, colors, and interactive options:

```{r}
nodes <- data.frame(id = 1:10, 
                    label = paste("Node", 1:10),
                    group = c("GrA", "GrB"),
                    value = 1:10, 
                    shape = c("square", "triangle", "box", "circle", "dot", "star",
                              "ellipse", "database", "text", "diamond"),
                    title = paste0("<p><b>", 1:10,"</b><br>Node !</p>"), 
                    color = c("darkred", "grey", "orange", "darkblue", "purple"),
                    shadow = c(FALSE, TRUE, FALSE, TRUE, TRUE)) 

edges <- data.frame(from = sample(1:10,8), to = sample(1:10, 8),
                    label = paste("Edge", 1:8), 
                    length = c(100,500), 
                    arrows = c("to", "from", "middle", "middle;to"),
                    dashes = c(TRUE, FALSE),  
                    title = paste("Edge", 1:8), 
                    smooth = c(FALSE, TRUE),  
                    shadow = c(FALSE, TRUE, FALSE, TRUE)) 

visNetwork(nodes = nodes, edges = edges) %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE)
```

## DataTables: Interactive Data Tables

The `DT` package renders interactive tables. Here, we display the `iris` dataset in a searchable and sortable format.

```{r}
datatable(data = iris, options = list(pageLength = 10), class = "cell-border stripe", editable = "cell")
```

# \( \LaTeX \): Mathematical Expressions

In RMarkdown, you can use LaTeX to format mathematical expressions. Here’s the formula for the three-parameter Generalized Pareto distribution (GPD):

\[
f(x) = \alpha^{-1}e^{-(1 - \kappa)y}, \quad y = 
\begin{cases} 
-\kappa^{-1}\log \Big[ 1 - \frac{\kappa(x - \xi)}{\alpha} \Big], & \kappa \neq 0,\\ 
\frac{(x - \xi)}{\alpha}, & \kappa = 0.
\end{cases}
\]

where:
- \( \xi \): location parameter
- \( \alpha > 0 \): scale parameter
- \( \kappa \): shape parameter, affecting the range of \( x \).

This setup provides a flexible base for using RMarkdown to demonstrate code, visualizations, and interactivity.
